---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Get started with projr

This guide shows you the minimal setup path to start using projr for reproducible research projects.

## Installation

```{r eval=FALSE}
# Once on CRAN:
# install.packages("projr")

# For now, install from GitHub:
remotes::install_github("SATVILab/projr")
```

## Minimal setup

The quickest way to start is to create the essential directories and run your first build:

```{r eval=FALSE}
library(projr)

# Create core directories
dir.create("_raw_data", showWarnings = FALSE)
dir.create("_output",   showWarnings = FALSE)

# Run your first build
projr_build()
```

This will:
- Clear `_output` directory
- Render any R Markdown or Quarto documents
- Version raw data and outputs
- Create a manifest linking inputs to outputs

## Full initialisation (optional)

For a complete project setup with metadata, Git, and GitHub:

### Option 1: Sensible defaults

```{r eval=FALSE}
projr_init()  # Uses sensible defaults
```

### Option 2: Step-by-step prompts

```{r eval=FALSE}
projr_init_prompt()  # Guides you through each option
```

### Option 3: Include extras by default

```{r eval=FALSE}
projr_init_full()  # Includes additional features
```

These initialisation functions set up:
- Project directories (`_raw_data`, `_output`, `_tmp`, `docs`)
- `_projr.yml` configuration file
- `DESCRIPTION` file with project metadata
- README template
- Git repository (optional)
- GitHub repository (optional)
- `.gitignore` and `.Rbuildignore` files
- renv for dependency management (optional)

## Development vs final builds

### Development builds: `projr_build_dev()`

Use this whilst iterating on your code:

```{r eval=FALSE}
projr_build_dev()
```

Development builds:
- Route documents and outputs to the cache directory (`_tmp/projr/v<version>/`)
- **Do not** touch your `_output` directory (unless `clear_output = "pre"`)
- **Do not** bump the version number
- **Do not** upload to archives
- Perfect for testing changes without overwriting your latest release

**Example**: Inspecting cached outputs

```{r eval=FALSE}
# After a dev build, find your outputs in:
# _tmp/projr/v0.1.0/output/
# _tmp/projr/v0.1.0/docs/

# You can specify which docs to render:
projr_build_dev("analysis.Rmd")
```

### Final builds: `projr_build()`

Use this when you're ready to create a new version:

```{r eval=FALSE}
# Patch version bump (0.1.0 -> 0.1.1)
projr_build()
# or explicitly:
projr_build_patch()

# Minor version bump (0.1.0 -> 0.2.0)
projr_build_minor()

# Major version bump (0.1.0 -> 1.0.0)
projr_build_major()
```

Final builds:
- Clear and populate `_output` directory
- Bump the version number
- Version all project components
- Create manifest linking inputs to outputs
- Archive to configured destinations (GitHub Releases, OSF, local)
- Commit and push to Git (if configured)

## Restoring on a new machine

Clone the repository and restore raw data:

```{r eval=FALSE}
projr_restore_repo("owner/repo")
```

This will:
- Clone the repository from GitHub
- Restore raw data from configured archives (GitHub Releases, OSF, or local)
- Set up the project structure
- (Optionally) restore renv dependencies with `renv::restore()`

## Understanding defaults

When you initialise a project, projr sets up these defaults:

### Directory labels

- `raw-data`: `_raw_data` - Your source data
- `cache`: `_tmp` - Temporary files and dev build outputs
- `output`: `_output` - Final outputs (figures, tables)
- `docs`: `docs` - Rendered documents (HTML, PDF)

### Auto-ignore rules

projr automatically configures:
- `.gitignore` to exclude large files and cache directories
- `.Rbuildignore` for R package development (if applicable)

### Manifest

After each build, projr creates a manifest (`_output/manifest.csv`) that records:
- Which version of raw data was used
- Which version of code was used
- Which outputs were generated
- SHA-256 checksums for verification

This ensures you can always trace outputs back to their exact inputs.

### Changelog and build log

- `NEWS.md` - Manual changelog for human-readable release notes
- `BUILDLOG.md` - Automated build log (configure via YAML)

## Next steps

- Read [How-to guides](how-to-guides.html) for specific tasks
- Understand [Concepts](concepts.html) behind projr's design
- Explore [Design](design.html) philosophy
- Browse the [Reference](../reference/index.html) for all functions
