---
title: "Streamlining Your Project Workflow with `projr`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2-understanding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup, warning = FALSE, message = FALSE}
library(projr)
```

# Understanding the `projr` Workflow

**What You'll Learn:**

* How `projr` structures your project for clarity and easy sharing.
* The core functions for building and refining your project.
* Simple steps to connect your project to GitHub.

## The Heart of `projr`: The `_projr.yml` File

This file is your project's command center.  Let's break down its key sections:

* **`directories`:**
   - **Purpose:** Defines where essential project components live.
   - **Key Folders:**
      - `raw-data`: Your untouched, original data.
      - `cache`: Temporary files created during development.
      - `output`: Final results you want to share (figures, tables, etc.).
      - `docs`: Rendered documents (HTML, PDF).
   - **Benefits:**
        - Organized structure that's easy to understand.
        - Simplifies sharing specific parts of your project.
        - `projr` can smartly manage and protect your outputs.

Use .path_get()` to find these folders. Example: .path_get("raw-data") ` will return the path to your `raw-data` directory.

* **`build`:**
   - **Purpose:** Automates tasks that happen every time you build your project.
   - **Key Actions**
        - **Version Control (Git):** Decides how Git commits are linked to project versions.
        - **Uploads:** Effortlessly shares data, outputs, and documents to GitHub, OSF, or local folders.

## Building Your Project: Key Functions

Let's look at the two functions that are the heart of building your project:

* **.build_dev()`**
   - **Purpose:** Your development buddy!  Renders your code documents, with results temporarily saved to the "cache" folder.
   - **Why Use It:**  Test changes safely before finalizing your outputs.

* **projr_build_output()`**
   - **Purpose:** The final production step. This saves results to their final locations *and* handles versioning, uploads, and other tasks you've set in `_projr.yml`.

## Notes on Git and GitHub

`projr` makes using Git easy, streamlining your version control and project sharing workflow. Here's how to connect your project to GitHub:

1. **Create a free GitHub account:** Sign up on GitHub: [https://github.com](https://github.com).
2. **GitHub Credentials:**  Follow the steps from Happy Git with R: [https://happygitwithr.com/https-pat#tldr](https://happygitwithr.com/https-pat#tldr) (See below for a quick summary).

**Generate a GitHub PAT:**

```{r, eval = FALSE}
if (!requireNamespace("usethis", quietly = TRUE)) {
  install.packages("usethis")
}
usethis::create_github_token()
```

Copy the generated token.

**Save Your Credentials:**

```{r, eval = FALSE}
gitcreds::gitcreds_set()
```

Paste your Git credentials when prompted.

## Next Steps

1. **Experiment:** Create a test project to practice using `projr`.
2. **Customize:** Tailor your `_projr.yml` file to match your specific project's needs.  Check out the `3-directories` and `4-destination` vignettes for more details!
3. **Integrate:** Make `projr` a natural part of your R workflow.

**Let's transform your projects from chaotic to streamlined!** 
