---
title: "Environment variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{environment-variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(projr)
```

## Project initialisation

The environment variables make custom setup of the project easier.

| Name | Value(s) | Action |
|------|----------|----------|--------|
| .PATH_YML` | `<path>` | Use file at `<path>` instead of default `_projr.yml` |
| .FIRST_NAME` | `<name>` | Offer `<name>` as first name option |
| .LAST_NAME` | `<name>` | Offer `<name>` as last name option |
| .EMAIL` | `<email>` | Offer `<email>` as email option |
| .GITHUB_USER_NAME` | `<username>` | Offer `<username>` as GitHub username option |


If set, .PATH_YML` specifies the default location for the `_projr.yml` file.
It may be overriden in .init` by setting the `yml_path_from` parameter.
If neither are specified, the default `_projr.yml` file is used.

The remaining environment variables are used to pre-populate the interactive
setup prompts.  If set, the user will be given these as options, but may
choose to manually enter an option instead.
Multiple options may be provided by placing a semi-colon between them, e.g.
"John;JohnnyBoy" yields two options: "John" and "JohnnyBoy".

## Project builds

During `projr` builds, `projr` mimicks the treatment of environment variables by `Quarto` projects.
Various files, prefixed with `_environment`, specify default values
for environment variables that are used if the variables are not already set.
Any variables not originally set are unset after the build.

Environment files are searched for in the following order:

- `_environment.local`
  - Intended to store user-specific, and possibly private, environment variables, such as access tokens.
  - This file should not be checked into Git, and is automatically ignored by `projr`.
- `_environment.<QUARTO_PROFILE>`, where `QUARTO_PROFILE` is the value of the
  `QUARTO_PROFILE` environment variable
    - Intended to store environment variables specific to a particular profile.
    - If `QUARTO_PROFILE` contains commas, then 
  the strings between commas are treated as separate profiles loaded in the order
  found in `QUARTO_PROFILE`.
- `_environment`
  - Intended to store environment variables common to all profiles.

In addition, if an `_environment.required` file is set, then the environment variables
specified inside are checked for existence.
If they are not set, warnings are printed (but the build is not aborted).
