---
title: "Environment variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{environment-variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(projr)
```

## Initialisation

The environment variables make custom setup of the project easier.

| Name | Value(s) | Action |
|------|----------|----------|--------|
| `PROJR_PATH_YML` | `<path>` | Use file at `<path>` instead of default `_projr.yml` |
| `PROJR_FIRST_NAME` | `<name>` | Offer `<name>` as first name option |
| `PROJR_LAST_NAME` | `<name>` | Offer `<name>` as last name option |
| `PROJR_EMAIL` | `<email>` | Offer `<email>` as email option |
| `PROJR_GITHUB_USER_NAME` | `<username>` | Offer `<username>` as GitHub username option |


If set, `PROJR_PATH_YML` specifies the default location for the `_projr.yml` file.
It may be overriden in `projr_init` by setting the `yml_path_from` parameter.
If neither are specified, the default `_projr.yml` file is used.

The remaining environment variables are used to pre-populate the interactive
setup prompts.  If set, the user will be given these as options, but may
choose to manually enter an option instead.
Multiple options may be provided by placing a semi-colon between them, e.g.
"John;JohnnyBoy" yields two options: "John" and "JohnnyBoy".

## 