---
title: "Project in Chaos? Get Organized with `projr`!"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(projr)
```

# Make Your R Projects Reproducible, Organized, and Shareable 

Let's be honest, organizing projects in R can feel like herding cats.  Code in one place, raw data scattered around, and don't even get started on trying to remember exactly which *versions* of everything worked together. That's where `projr` steps in to streamline your workflow.

## What is `projr`?

Think of `projr` as your R project's personal assistant. It helps you:

* **Set up a consistent structure:** No more "Where should this file go?" dilemmas. `projr` guides you with a clear project layout.
* **Track changes:** Code changes? Data updates? `projr` automatically versions your project, so you have a crystal-clear history.
* **Collaborate seamlessly:** Sharing your work is a breeze with built-in tools for version control (we'll take care of the tricky Git stuff for you!).

## Why should I care?

1. **Reproducibility:** Anyone (including your future self!) can easily rerun your analysis with all the right ingredients.
2. **Organization:**  A well-organized project is a happy project.  Find what you need, when you need it.
3. **Collaboration:** Work smoothly with others, knowing your project's history and files are easily shared.

## How does it work? (The Super-Simplified Version)

1. **Start a new project:** .init()` sets the stage with folders for your data, code, and results. It even connects to GitHub if you want.
2. **Build your project:** projr_build_output()` runs your code, keeps track of versions, and neatly packages everything up.

That's the gist, but `projr` can do much more. It can automatically share your project on platforms like GitHub or store it locally – all customizable with a simple configuration file.

## Ready to get started? 

**1. Install `projr`:**

```{r, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("SATVILab/projr")
library(projr)
```

**2.  Open R in your desired project folder.**  (If you're using RStudio, it's easiest to start a new project right inside the folder.)

**3. Initialize your project:**

```r
projr_init() 
```

Follow the prompts, and `projr` will take care of the rest!

## Building Your Project

Now it's time to make your project come alive!

* **Test it out:** Use .build_dev()` to run your code and see how your results look. This is a great way to catch errors as you go.

* **Ready to share?** Use projr_build_output()` to create the final version of your project. This will automatically version everything, package it up nicely, and even share it on platforms like GitHub if you've set that up.

## Where can I save my stuff?

`projr` has a special configuration file (`_projr.yml`) that tells it where to store your data, results, and the beautiful documents you create.  You can customize this to fit your workflow perfectly.  

**Sound good? Let's dive into the details in the next vignettes!** 
